---
title: "Introduction to fairGATE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to fairGATE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  error = TRUE 
)
```

# 1. Introduction

The **fairGATE** package provides a complete pipeline for training and evaluating a Gated Neural Network (GNN) designed to mitigate demographic bias in predictive modelling. The package implements a fairness-aware GNN that uses a custom loss function to enforce the *Equalized Odds* fairness criterion by minimising the variance in True Positive and False Positive Rates across subgroups.

This vignette demonstrates the full workflow using the GENDEP dataset to predict antidepressant response, focusing on fairness across gender subgroups.

# 2. The fairGATE Workflow

The package is built around a logical sequence of functions:

-   `prepare_data()`: Cleans and prepares the input data.\
-   `train_gnn()`: Trains the Gated Neural Network.\
-   `analyse_gnn_results()`: Conducts performance and gate analysis.\
-   `analyse_experts()`: Performs analysis of expert specialisation.\
-   `plot_sankey()`: Visualises the model's patient routing behaviour.

## 2.1. Loading Data

First, we load the necessary libraries and the dataset.

We use a small in-package sample of the UCI Adult dataset. Outcome is \*\*income\*\* (1 = \>50K, 0 = ≤50K); protected attribute is \*\*sex\*\*.

```{r}
library(fairGATE)
library(dplyr)
library(readxl)

# A user would load their own data here.
data("adult_sample", package = "fairGATE")

adult <- adult_sample %>%
  mutate(
    across(where(is.character), ~ trimws(.x)),
    income = as.integer(income)
  )
```

## 2.2. Step 1: Prepare the Data

We use `prepare_data()` to process the raw data for the Male/Female analysis.

```{r}

# Dropping unwanted cols (i.e. numeric cols and those with high multicolinearity)
cols_to_drop <- c("subjectid", "Row.names")

#  Ensure to perform other preprocessing steps such as one-hot endoing etc

# Fully prepared data goes here

prepared <- fairGATE::prepare_data(
  data          = adult,
  outcome_var   = "income",
  group_var     = "sex",
  cols_to_remove= cols_to_drop
)
```

## 2.3. Train a small demo model

```{r train-demo, results='hide', message=FALSE, warning=FALSE}
# Train a small Gated Neural Network
trained_model <- fairGATE::train_gnn(
  prepared_data = prepared,
  run_tuning    = FALSE,     # skip tuning for speed
  best_params   = list(
    lr = 0.01,
    hidden_dim = 16,
    dropout_rate = 0.1,
    lambda = 0.0,
    temperature = 1.0
  ),
  num_repeats   = 2,         # very short repeated split
  epochs        = 20,        # fast CRAN-safe runtime
  verbose       = FALSE
)
```

## 2.4. Step 3: Analyse Basic Performance and Gates

With the results loaded, we run `analyse_gnn_results()` to generate all the standard performance plots and gate analyses.

```{r}

# Run basic analysis
basic_analyses <- analyse_gnn_results(
  gnn_results = trained_model,
  prepared_data = prepared_data_gender
)

# --- View all plots from the basic analysis ---
cat("## ROC Curve\n")
print(basic_analyses$roc_plot)

cat("\n## Calibration Plot\n")
print(basic_analyses$calibration_plot)

cat("\n## Gate Weight Distribution\n")
print(basic_analyses$gate_density_plot)

cat("\n## Gate Entropy Distribution\n")
print(basic_analyses$entropy_density_plot)
```

## 2.5. Step 4: Analyse Expert Specialisation

Now, we use `analyse_experts()` to investigate how the different expert networks have specialised their learning.

```{r}
# --- View all results from the expert analysis ---
cat("\n## Feature Importance: Female vs. Male\n")
# This table shows the features with the biggest difference in importance between the two experts.
print(head(expert_analyses$pairwise_differences$Female_vs_Male))

cat("\n## Feature Importance Difference Plot\n")
# This plot visualises the differences from the table above.
print(expert_analyses$difference_plot)
```

## 2.6. Step 5: Visualise Patient Routing with a Sankey Plot

Finally, we use `plot_sankey()` to create the key visualisation from the research paper, showing how patients are routed through the model.

```{r}
# Generate and print the Sankey plot
sankey_diagram <- plot_sankey(
    raw_data = raw_data,
    gnn_results = gnn_results,
    expert_results = expert_analyses,
    group_mappings = label_mappings_gender,
    group_var = "sex"
)

print(sankey_diagram)
```
